\texttt{davos}: a Python package ``smuggler'' for constructing lightweight reproducible notebooks

### 2022-11-20 paper/main.tex
Section -> Sec. in parens

The \texttt{davos} package consists of two interdependent subpackages
(see Fig.~\ref{fig:package-structure}). The first,
\texttt{davos.core}, comprises a set of modules that
implement the bulk of the package's core
functionality, including pipelines for installing and validating
packages, custom parsers for the \texttt{smuggle} statement (see
Section~\ref{subsec:smuggle}) and onion comment (see
Section~\ref{subsec:onion}), and a runtime interface for configuring
\texttt{davos}'s behavior (see Section~\ref{subsec:config}). However,
certain critical aspects of this
functionality require (often substantially) different implementations
depending on properties of the notebook environment in which
\texttt{davos} is used (e.g., whether the frontend is provided by
Jupyter or Google Colaboratory, or which version of
IPython~\cite{PereGran07} is used by the notebook kernel). To deal
with this, environment-dependent parts of core features and behaviors
are isolated and abstracted to ``helper functions'' in the
\texttt{davos.implementations} subpackage. This second subpackage
defines multiple, interchangeable versions of each helper function,
organized into modules by the conditions that trigger their use. At
runtime, \texttt{davos} detects various features in the notebook
environment and selectively imports a single version of each helper
function into the top-level \texttt{davos.implementations} namespace,
allowing \texttt{davos.core} modules to access the proper
implementations for the current notebook environment in a single,
consistent location. An additional benefit of this design is that it
allows maintainers, developers, and users to extend \texttt{davos} to
support new, updated, or custom notebook variants by creating new
\texttt{davos.implementations} modules that define their own versions
of each helper function, modified from existing implementations as
needed.

@@

Functionally, importing \texttt{davos} in an IPython notebook enables
an additional Python keyword: ``\texttt{smuggle}'' (see
Section~\ref{subsec:implementation} for details on how this works).
The \texttt{smuggle} keyword-like object can be used as a drop-in
replacement for Python's built-in \texttt{import} keyword to load
packages, modules, and other objects into the current namespace.
However, whereas \texttt{import} will fail if the requested package is
not installed locally, \texttt{smuggle} statements can handle missing
packages on the fly.  If a smuggled package does not exist in the
local environment, \texttt{davos} will download and install it automatically,
expose its contents to Python's \texttt{import} machinery, and load it
into the namespace for immediate use.

@@

For greater control over the behavior of \texttt{smuggle} statements,
\texttt{davos} defines an additional construct called the ``onion
comment.'' An onion comment is a special type of inline comment that
may be placed on a line containing a \texttt{smuggle} statement to
customize how \texttt{davos} searches for the smuggled package locally
and, if necessary, downloads and installs it. Onion comments follow a
simple format based on the ``type comment'' syntax introduced in PEP
484~\cite{vanREtal14}, and are designed to make managing packages with
\texttt{davos} intuitive and familiar. To construct an onion comment,
users provide the name of the installer program (e.g., \texttt{pip})
and the same arguments one would use to manually install the package
as desired via the command line:
\begin{center}
\includegraphics[width=0.9\textwidth]{figs/snippet1}
\end{center}
Occasionally, a package's distribution name (i.e., the name used when
installing it) may differ from its top-level module name (i.e., the
name used when importing it). In such cases, an onion comment may be
used to ensure that \texttt{davos} installs the proper package if it
cannot be found locally:
\begin{center}
\includegraphics[width=0.9\textwidth]{figs/snippet2}
\end{center}
Because onion comments may be constructed to specify any aspect of the
installer's behavior, they provide a mechanism for precisely
controlling how, where, and when smuggled packages are installed.
Critically, if an onion comment includes a version
specifier~\cite{CoghStuf13}, \texttt{davos} will ensure that the
version of the package loaded into the notebook matches the specific
version requested, or satisfies the given version constraints. If the
smuggled package exists locally, \texttt{davos} will extract its
version information from its metadata and compare it to the specifier
provided. If the two are incompatible (or no local installation is
found), \texttt{davos} will download, install, and load a suitable
version of the package instead:
\begin{center}
\includegraphics[width=0.9\textwidth]{figs/snippet3}
\end{center}
Onion comments can also be used to smuggle specific VCS references
(e.g., Git~\cite{TorvHama05} branches, commits, tags, etc.):
\begin{center}
\includegraphics[width=0.9\textwidth]{figs/snippet4}
\end{center}
\texttt{davos} processes onion comments internally before forwarding
arguments to the installer program. In addition to preventing onion
comments from being used as a vehicle for shell injection attacks,
this enables \texttt{davos} to adapt its behavior based on how particular
flags will affect the behavior of the installer program. For example, if an
onion comment contains either the \texttt{-I}/\texttt{--ignore-installed},
\texttt{-U}/\texttt{--upgrade}, or \texttt{--force-reinstall} flag,
\texttt{davos} will not bother checking for a local copy of the smuggled
package before installing a new one:
\begin{center}
\includegraphics[width=0.9\textwidth]{figs/snippet5}
\end{center}
Similarly, the \texttt{--no-input} flag will temporarily enable
\texttt{davos}'s non-interactive mode (see
Section~\ref{subsec:config}), and installing a smuggled package into a
custom directory (\texttt{<dir>}) using the \texttt{--target <dir>}
flag will cause \texttt{davos} to prepend \texttt{<dir>} to the module
search path (i.e., \texttt{sys.path}), if necessary, so the package
can be imported.

@@

After installing and importing \texttt{davos} (lines 1--2), we first smuggle two
utilities for interacting with local files in the code below. The
\texttt{smuggle} statement in line 4 loads the \texttt{is\_file()}
function from the Python standard library's \texttt{os.path}
module. Standard library modules are included with all Python
distributions, so this line is functionally equivalent to an
\texttt{import} statement and does not need or benefit from an onion
comment. Line 5 loads the \texttt{joblib} package~\cite{Varo10},
installing it first, if necessary. Since \texttt{joblib}'s I/O
interface has historically remained stable and backwards-compatible
across releases, requiring that users have a particular exact version
installed would likely be unnecessarily restrictive. However, a
\textit{future} release might introduce some breaking change.  The
onion comment in line 5 helps ensure the analysis notebook continues
to run properly in the future, by limiting allowable versions to those
already released when the code was written:
\begin{center}
\includegraphics[width=0.9\textwidth]{figs/example1}
\end{center}
Line 7 then uses the \texttt{davos.config} object to enable
\texttt{davos}'s \texttt{auto\_rerun} option before smuggling the next
two packages: \texttt{NumPy}~\cite{HarrEtal20} and
\texttt{pandas}. Because these packages rely heavily on custom C data
types, loading the particular versions from the onion comments may
require restarting the notebook kernel if different versions had been previously
imported during the same interpreter session (see
Section~\ref{subsec:config}).
\begin{center}
\includegraphics[width=0.9\textwidth]{figs/example2}
\end{center}
Setting the \texttt{auto\_rerun} attribute to \texttt{True} also
benefits the (potential) installation of \texttt{pandas} on the next
lines:
\begin{center}
\includegraphics[width=0.9\textwidth]{figs/example3}
\end{center}
If we suppose that the data contained in \texttt{data-old.pkl} is
stored in a \texttt{Panel} object (i.e., created in a version of
\texttt{pandas} prior to 0.25.0), then we would not be able to load in
that object with newer versions of \texttt{pandas}.  Line 11 ensures
that an older version of \texttt{pandas} will be imported, enabling
the data to be read in (and, in line 13, saved out to a \texttt{.csv}
file, which is compatible with newer versions of \texttt{pandas}).

@@

Our work also has several more subtle ``advanced'' use cases and
potential impacts. Whereas Python's built-in \texttt{import} statement
is agnostic to packages' version information, \texttt{smuggle}
statements (when combined with onion comments) are version-sensitive.
And because onion comments are parsed at runtime, required packages
and their specified versions are installed in a just-in-time
manner. Thus, it is possible in most cases to \texttt{smuggle} a
specific package version or revision even if a different version has
already been loaded. This enables more complex uses that take
advantage of multiple versions of a package within a single
interpreter session (e.g., see
Section~\ref{sec:illustrative-example}). This could be useful in cases
where specific features are added or removed from a package across
different versions, or in comparing the performance or functionality
of particular features across different versions of the same package.

>>>>

The \texttt{davos} package consists of two interdependent subpackages
(see Fig.~\ref{fig:package-structure}). The first,
\texttt{davos.core}, comprises a set of modules that
implement the bulk of the package's core
functionality, including pipelines for installing and validating
packages, custom parsers for the \texttt{smuggle} statement (see
Sec.~\ref{subsec:smuggle}) and onion comment (see
Sec.~\ref{subsec:onion}), and a runtime interface for configuring
\texttt{davos}'s behavior (see Sec.~\ref{subsec:config}). However,
certain critical aspects of this
functionality require (often substantially) different implementations
depending on properties of the notebook environment in which
\texttt{davos} is used (e.g., whether the frontend is provided by
Jupyter or Google Colaboratory, or which version of
IPython~\cite{PereGran07} is used by the notebook kernel). To deal
with this, environment-dependent parts of core features and behaviors
are isolated and abstracted to ``helper functions'' in the
\texttt{davos.implementations} subpackage. This second subpackage
defines multiple, interchangeable versions of each helper function,
organized into modules by the conditions that trigger their use. At
runtime, \texttt{davos} detects various features in the notebook
environment and selectively imports a single version of each helper
function into the top-level \texttt{davos.implementations} namespace,
allowing \texttt{davos.core} modules to access the proper
implementations for the current notebook environment in a single,
consistent location. An additional benefit of this design is that it
allows maintainers, developers, and users to extend \texttt{davos} to
support new, updated, or custom notebook variants by creating new
\texttt{davos.implementations} modules that define their own versions
of each helper function, modified from existing implementations as
needed.

@@

Functionally, importing \texttt{davos} in an IPython notebook enables
an additional Python keyword: ``\texttt{smuggle}'' (see
Sec.~\ref{subsec:implementation} for details on how this works).
The \texttt{smuggle} keyword-like object can be used as a drop-in
replacement for Python's built-in \texttt{import} keyword to load
packages, modules, and other objects into the current namespace.
However, whereas \texttt{import} will fail if the requested package is
not installed locally, \texttt{smuggle} statements can handle missing
packages on the fly.  If a smuggled package does not exist in the
local environment, \texttt{davos} will download and install it automatically,
expose its contents to Python's \texttt{import} machinery, and load it
into the namespace for immediate use.

@@

For greater control over the behavior of \texttt{smuggle} statements,
\texttt{davos} defines an additional construct called the ``onion
comment.'' An onion comment is a special type of inline comment that
may be placed on a line containing a \texttt{smuggle} statement to
customize how \texttt{davos} searches for the smuggled package locally
and, if necessary, downloads and installs it. Onion comments follow a
simple format based on the ``type comment'' syntax introduced in PEP
484~\cite{vanREtal14}, and are designed to make managing packages with
\texttt{davos} intuitive and familiar. To construct an onion comment,
users provide the name of the installer program (e.g., \texttt{pip})
and the same arguments one would use to manually install the package
as desired via the command line:
\begin{center}
\includegraphics[width=0.9\textwidth]{figs/snippet1}
\end{center}
Occasionally, a package's distribution name (i.e., the name used when
installing it) may differ from its top-level module name (i.e., the
name used when importing it). In such cases, an onion comment may be
used to ensure that \texttt{davos} installs the proper package if it
cannot be found locally:
\begin{center}
\includegraphics[width=0.9\textwidth]{figs/snippet2}
\end{center}
Because onion comments may be constructed to specify any aspect of the
installer's behavior, they provide a mechanism for precisely
controlling how, where, and when smuggled packages are installed.
Critically, if an onion comment includes a version
specifier~\cite{CoghStuf13}, \texttt{davos} will ensure that the
version of the package loaded into the notebook matches the specific
version requested, or satisfies the given version constraints. If the
smuggled package exists locally, \texttt{davos} will extract its
version information from its metadata and compare it to the specifier
provided. If the two are incompatible (or no local installation is
found), \texttt{davos} will download, install, and load a suitable
version of the package instead:
\begin{center}
\includegraphics[width=0.9\textwidth]{figs/snippet3}
\end{center}
Onion comments can also be used to smuggle specific VCS references
(e.g., Git~\cite{TorvHama05} branches, commits, tags, etc.):
\begin{center}
\includegraphics[width=0.9\textwidth]{figs/snippet4}
\end{center}
\texttt{davos} processes onion comments internally before forwarding
arguments to the installer program. In addition to preventing onion
comments from being used as a vehicle for shell injection attacks,
this enables \texttt{davos} to adapt its behavior based on how particular
flags will affect the behavior of the installer program. For example, if an
onion comment contains either the \texttt{-I}/\texttt{--ignore-installed},
\texttt{-U}/\texttt{--upgrade}, or \texttt{--force-reinstall} flag,
\texttt{davos} will not bother checking for a local copy of the smuggled
package before installing a new one:
\begin{center}
\includegraphics[width=0.9\textwidth]{figs/snippet5}
\end{center}
Similarly, the \texttt{--no-input} flag will temporarily enable
\texttt{davos}'s non-interactive mode (see
Sec.~\ref{subsec:config}), and installing a smuggled package into a
custom directory (\texttt{<dir>}) using the \texttt{--target <dir>}
flag will cause \texttt{davos} to prepend \texttt{<dir>} to the module
search path (i.e., \texttt{sys.path}), if necessary, so the package
can be imported.

@@

After installing and importing \texttt{davos} (lines 1--2), we first smuggle two
utilities for interacting with local files in the code below. The
\texttt{smuggle} statement in line 4 loads the \texttt{is\_file()}
function from the Python standard library's \texttt{os.path}
module. Standard library modules are included with all Python
distributions, so this line is functionally equivalent to an
\texttt{import} statement and does not need or benefit from an onion
comment. Line 5 loads the \texttt{joblib} package~\cite{Varo10},
installing it first, if necessary. Since \texttt{joblib}'s I/O
interface has historically remained stable and backwards-compatible
across releases, requiring that users have a particular exact version
installed would likely be unnecessarily restrictive. However, a
\textit{future} release might introduce some breaking change.  The
onion comment in line 5 helps ensure the analysis notebook continues
to run properly in the future, by limiting allowable versions to those
already released when the code was written:
\begin{center}
\includegraphics[width=0.9\textwidth]{figs/example1}
\end{center}
Line 7 then uses the \texttt{davos.config} object to enable
\texttt{davos}'s \texttt{auto\_rerun} option before smuggling the next
two packages: \texttt{NumPy}~\cite{HarrEtal20} and
\texttt{pandas}. Because these packages rely heavily on custom C data
types, loading the particular versions from the onion comments may
require restarting the notebook kernel if different versions had been previously
imported during the same interpreter session (see
Sec.~\ref{subsec:config}).
\begin{center}
\includegraphics[width=0.9\textwidth]{figs/example2}
\end{center}
Setting the \texttt{auto\_rerun} attribute to \texttt{True} also
benefits the (potential) installation of \texttt{pandas} on the next
lines:
\begin{center}
\includegraphics[width=0.9\textwidth]{figs/example3}
\end{center}
If we suppose that the data contained in \texttt{data-old.pkl} is
stored in a \texttt{Panel} object (i.e., created in a version of
\texttt{pandas} prior to 0.25.0), then we would not be able to load in
that object with newer versions of \texttt{pandas}.  Line 11 ensures
that an older version of \texttt{pandas} will be imported, enabling
the data to be read in (and, in line 13, saved out to a \texttt{.csv}
file, which is compatible with newer versions of \texttt{pandas}).

@@

Our work also has several more subtle ``advanced'' use cases and
potential impacts. Whereas Python's built-in \texttt{import} statement
is agnostic to packages' version information, \texttt{smuggle}
statements (when combined with onion comments) are version-sensitive.
And because onion comments are parsed at runtime, required packages
and their specified versions are installed in a just-in-time
manner. Thus, it is possible in most cases to \texttt{smuggle} a
specific package version or revision even if a different version has
already been loaded. This enables more complex uses that take
advantage of multiple versions of a package within a single
interpreter session (e.g., see
Sec.~\ref{sec:illustrative-example} and Fig.~\ref{fig:illustrative-example}). This could be useful in cases
where specific features are added or removed from a package across
different versions, or in comparing the performance or functionality
of particular features across different versions of the same package.
